### 2020-02-21 Writing/disparity_review.tex
Updated until phylo methods

\section{Data and disparity}

@@

\subsection{From data to morphospace}

The concept of what the morphospace \textit{is} can be different between authors and studies but generally used in the common (i.e. non-mathematical) sense of a space containing all possible morphologies.
For the purpose of this review, we use a generalised version of the definition of \cite{mitteroecker2009concept} where the morphospace is a \textit{p}-dimensional space (or Q-space) containing \textit{n} points \citep{mitteroecker2009concept}.
In other words, mathematically, the morphospace is the mathematical object that contains the trait values (\textit{p}, the axes of the space) for all the observations (\textit{n}, the specimens); or, biologically, the morphospace contains all the observed combination of traits.
We generalise this definition to any linear transformation ($V \to W$) of this space that retains the biological definition, i.e. the morphospace is the mathematical object that contains the observed combinations of traits or their transformation.
Note that here the morphospace does not need to be Euclidean (although this is often desired or assumed - see discussion below). %link
If the morphospace is transformed mathematically (i.e. the dataset is transformed using logs, ordinations, distance matrices, etc...) it remains the morphospace as long as all the observation and traits remain in the datatset.

@@

\section{Disparity analysis methods}

@@

\subsection{(a) To ordinate or not to ordinate - that is the (multidimensional) question}

@@

One of the reason why ordinations techniques are common in disparity analysis is that they make it easier for researchers to comprehend patterns in two or three spatial dimensions at the time which can be more intuitively captured than through disparity indices (see section below). %@@@ LINK
Additionally, after ordinating the data, it is possible to focus on just a subset of axes of the morphospace (i.e. selecting only those axes that describe the majority of the variation in the dataset - e.g. 95\%).
In the case of geometric morphometric data, ordination is particularly useful as it conserves the mathematical properties of the data while efficiently reducing the dimensions \citep{Legendre2012-va,dryden2016statistical}.
In practice, this can allow to interpret only the major axis of a highly dimensional dataset as major gradients of biological variation \citep[e.g. the elongation and flattening of birds beaks;][]{Cooney2017-ly}.

@@

Finally, categorical data are a good deal more problematic, since the characters themselves are invariably non-equivalent, non-independent, and mostly non-Euclidean and the distribution of the variance is usually normal (i.e. contrary to a PCA, the first few axis do not encompass most variance of the dataset).
Such non-Euclidean spaces often have non-intuitive properties, for example, straight lines viewed in bivariate plots of some two selected dimensions are not actually straight and, even less intuitively, distances are non-metric \citep[i.e. the distance between A and B is not equal to the distance between B and A][]{Gerber2014-ol}.
And last but not least reason, In many cases, ordination might not be necessary.
For example, if a metric (e.g. sum or ranges of variance) used to characterise disparity can use all of the data, it is not necessary to calculate it on the ordinated dataset \citep{Close2015-qi}.
For all of these reasons, multidimensional data analyses should not be ordinated automatically, and careful consideration should be given to whether the aim of the study can be achieved without ordination \citep{lloyd2016,lloyd2018}.

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.9\textwidth]{Figures/figure_space.pdf}
\caption{
    Morphospaces: different mathematical representations of a morphospace.
    A trait matrix can be transformed into a distance matrix \citep[e.g in][]{Close2015-qi} or an ordinated matrix \citep[e.g. in][]{Brusatte2008-vx}.
    Here we consider all these matrices as being \textit{morphospaces}, i.e. objects containing all the combinations of traits and observations (albeit transformed differently).
    Visualisation: different  ways to represent the morphospace in 2D.
    Visualisations can use either trait plots (directly from the trait matrix); ``flattening'' ordination of the data (e.g. using multidimensional scaling forcing all the variance to be contained in two dimensions)
    % SG: No, the flattening will not force all the variance to be contained in 2D. First because NMDS is non metric and variance is therefore not preserved as such. Second, because there is a stress associated with the flattening that correlates with the degree of distortion imposed on the data to enforce the 2D representation. Some information (variance) is thus necessarily lost.
    ; or ordination axis plots (directly from the ordinated matrix).
    % ES: I feel like somewhere we should say that in order for the spaces to be Euclidean, the axes must be on same scale. My little bugbear.
    % SG response: I think there are more conditions to be Euclidean (inner product defining the  Euclidean norm..). But surely if axes have different scales and units, it will be difficult to define a distance function and the space won't be metric.
    % @@@ Change references
}
\label{Fig:morphospace}
\end{figure}

% @@@ Change to indices throughout
\subsection{(b) Measuring disparity}

Most disparity datasets are multidimensional and, consequently, a large component of any disparity analysis involves considering how to extract a meaningful (i.e. interpretable) summary of disparity.
This is usually achieved with a disparity metric or index \citep{Hopkins2017-cf}.
As with any summary of multidimensional data, disparity metrics will reflect only some aspects of the morphological variation, never its whole complexity %metrics paper. CITE moms
It is therefore often beneficial to use more than one metric to summarise different aspects of variation, guided by the aim of the study.

When considering only one dimension, disparity metrics can be used to reflect the spread of the distribution (e.g. the range, quantiles or variance) or its central tendency (i.e., mean, median or mode).
Among these metrics, some will have more attractive properties than others, such as sensitivity to outliers. Range and mean are highly sensitive, whereas quantiles, variance and median are less so, making them more or less appropriate for different questions.
For example, if we want to characterise the extent of morphospace occupied by a group (e.g. does group A occupy as much space as group B?), metrics related to the spread of the group in the morphospace are most appropriate (e.g. volume
\citealt{Diaz2016-mr}, distance from the centroid \citealt{Hopkins2017-cf, Finlay2015-ft}, variance and range \citealt{Brusatte2008-fx}).
Conversely, if we wish to describe the ``position''
% SG: Note that conceptually and technically,  disparity is about variation. So information on position is another aspect of morphospace occupation. but it is distinct from the meaning of  disparity as originally intended.
 of a group in a morphospace (e.g. does group A occupy the same space as group B?), metrics related to the distance between the elements within a group and a fixed point in the morphospace are most appropriate. %@@@ CITE moms
Finally, if we aim to characterise the density of morphospace occupation (e.g. is group A more closely packed than group B?) metrics related to the pairwise distances between elements will be most appropriate (e.g. nearest neighbour distance, pairwise distances, etc. \citealt{Close2015-qi}).
%ES: this para is probably most useful to refer to for the ordinate or not part

In addition to considering what properties of disparity metrics should capture, it is important to also consider the mathematical properties of the metrics and their associated caveats \citep{Wills2001-wh, Ciampaglio2001-iz}.
For example, measuring the full sum of the variance in each dimension of the space does not require we add the covariance between the axes in an ordinated space using a PCA.
% SG: Not very clear to me
However, this is not true of other mathematical spaces or when not all dimensions or elements are considered, even in a PCA \citep{Legendre2012-va}.

Furthermore, multidimensional spaces also have some counter-intuitive properties that need to be considered such as the ``curse of dimensionality'' \citep{Bellman1966-mc}.
In spaces with some axis of variance lower than 1, product-based metrics used as proxies of volumes (e.g. product of ranges, hypervolume, hypercube, etc.) can tend towards zero fairly quickly for spaces with even a modest number of dimensions \citep{Bellman1966-mc}. %Donoho 2000
Some other types of metrics are also extremely sensitive to outliers and can be biased easily by sample size, for example range \citep{Butler2012-tr}
%SG: Or even better, twenty years earlier, Foote (1992), who orginally described the problem in the context of disparity and suggested a rarefaction approach.
or convex hull based metrics \citep{Butler2012-tr, Jackson2011-kq}.

\subsection{(c) Testing hypotheses on the evolution of
disparity}

No matter which disparity metrics have been calculated, the research question must be framed in an appropriate statistical context.
The multidimensional statistical toolkit for ecology and evolution has been greatly expanded in recent years \citep{Adams2018-mg}
%AG: there are others that should be mentioned - mvmorph for example, which is especially important given that there are extremely different approaches to multivariate data in these different packages
, but some of these advances have yet to be implemented in disparity analyses. Instead, hypothesis testing has mostly been confined to a small set of well-established methods.
One commonly used test is the non-parametric permutation analysis of variance \citep{Anderson2001-qb, Anderson2013-zt}, an analysis of variance (ANOVA) of the pairwise distances between different groups.
Although statistically valid, this test is not always directly related to the hypothesis under question.
For example, PERMANOVA tests whether two groups share the same variance/covariance in a ``distance-space''.
% SG: What is meant by distance-sapce and in what way is it distinguished from the spaces discussed above?
This is not the same as testing whether the two groups overlap in morphospace.
%SG: You're sure of this?
Statistical tests should be employed that are tailored to the question at hand, rather than simply following common practices.

It is also important to consider which data should be subjected to a statistical test.
% AG: It would be good to also note here that data should be comparable in terms of having a common reference frame - e.g. comparing disparities across elements with different numbers of landmarks or different landmark configurations that haven't been/can't be superimposed together is problematic but is popping up in the disparity literature
For example, in morphological disparity analysis, especially for palaeobiological questions, data are often bootstrapped.
This has two advantages: (i) when the disparity metric is unidimensional (e.g. the sum of variances), bootstrapping the data generates a distribution of the metric that can be analysed using the vast statistical toolkit available for comparing distributions; (ii) when data are scarce, bootstrapping the data allows users to introduce variance, rendering the test less sensitive to outliers.
However, bootstrapped data are pseudoreplicates and thus non-independent.
This violates the assumptions of most parametric statistical tests.
Furthermore, the number of bootstrap pseudoreplicates will inevitably increase the false positive rate (Type I error).
% @@@ check positive/negative type
These factors are often ignored in disparity analyses.
% VW: Surely there's a reference for this?
% MH: I might be wrong about this, but my impression is that CIs generated from bootstrapping are compared visually (like to assess significance in changes in disparity over time), but that statistical tests are not commonly applied.

%TG: Add section d) within c)
\subsection{(d) Phylogenetic autocorrelation}

As with all comparative datasets, the data used in disparity analyses are not independent because close relatives will tend to have more similar morphologies than more distant relatives \citep{Harvey1998-xg}.
Thus, for many disparity analyses, phylogenetic relationships should be
taken into account.
% SG: Why?
However, it has been noted that some popular phylogenetic correction methods (like pPCA) can be inappropriate
% SG: In what way?
if incorrect assumptions are made about the data \citep{Uyeda2015}.
% @@@ref: https://academic.oup.com/sysbio/article/64/4/677/1649888
% SG: Which assumption? This part (e) seems a bit vague to me
% AG: as long as you are using most or ideally all of the variance in a dataset, the rotation of pPCA should have little to no impact on disparity analyses. polly et al 2013 i think
provide a thorough review of multivariate phylogenetic comparative methods, and so we do not consider them further here.
% AG: Given this is given quite short treatment, i think if you need room, you could cut it altogether or just have a sentence in another section stating this even more briefly

%TG: Add section e) within c)???
\subsection{(e) Ancestral state estimation in disparity-through-time analyses}

@@

\section{Disparity analyses for the future}

@@

\subsection{Appendix: morphospace definition in this paper}

>>>>

\section{Data and disparity}
\label{section:data}

@@

\subsection{From data to morphospace}
\label{section:morphospace}

% TG: The whole argument here is the same as for logging data in biology: we conserve the same properties of the biological observation (e.g. a distribution of body mass) but we modify it linearly to make it easier to analyse (e.g. the log(body mass)). In my opinion same goes with a data matrix: a distance matrix or a PCA of the observed trait matrix should no modify the biological properties as long as it it changed linearly (if you start arbitrarily removing axis of the PCA or points in the distance matrix, this results in something different). Therefore, morphospace = collected data matrix (any any linear transformation thereof).

The concept of what the morphospace \textit{is} can be different between authors and studies but generally used in the common (i.e. non-mathematical) sense of a space containing all possible morphologies.
For the purpose of this review, we use a generalised version of the definition of \cite{mitteroecker2009concept} where the morphospace is a \textit{p}-dimensional space (or Q-space) containing \textit{n} points \citep{mitteroecker2009concept}.
In other words, mathematically, the morphospace is the mathematical object that contains the trait values (\textit{p}, the axes of the space) for all the observations (\textit{n}, the specimens); or, biologically, the morphospace contains all the observed combination of traits.
We generalise this definition to any linear transformation ($V \to W$) of this space that retains the biological definition, i.e. the morphospace is the mathematical object that contains the observed combinations of traits or their transformation (see appendix \ref{section:appendixmorphospace}).
Note that here the morphospace does not need to be Euclidean (although this is often desired or assumed - see section \ref{section:ordination} below).
If the morphospace is transformed mathematically (i.e. the dataset is transformed using logs, ordinations, distance matrices, etc...) it remains the morphospace as long as all the observation and traits remain in the datatset.

@@

\section{Disparity analysis methods}
\label{section:methods}

@@

\subsection{(a) To ordinate or not to ordinate - that is the (multidimensional) question}
\label{section:ordination}

@@

One of the reason why ordinations techniques are common in disparity analysis is that they make it easier for researchers to comprehend patterns in two or three spatial dimensions at the time which can be more intuitively captured than through disparity indices (see section \ref{section:metrics} below).
Additionally, after ordinating the data, it is possible to focus on just a subset of axes of the morphospace (i.e. selecting only those axes that describe the majority of the variation in the dataset - e.g. 95\%).
In the case of geometric morphometric data, ordination is particularly useful as it conserves the mathematical properties of the data while efficiently reducing the dimensions \citep{Legendre2012-va,dryden2016statistical}.
In practice, this can allow to interpret only the major axis of a highly dimensional dataset as major gradients of biological variation \citep[e.g. the elongation and flattening of birds beaks;][]{Cooney2017-ly}.

@@

Finally, categorical data are a good deal more problematic, since the characters themselves are invariably non-equivalent, non-independent, and mostly non-Euclidean and the distribution of the variance is usually normal (i.e. contrary to a PCA, the first few axis do not encompass most variance of the dataset).
Such non-Euclidean spaces often have non-intuitive properties, for example, straight lines viewed in bivariate plots of some two selected dimensions are not actually straight and, even less intuitively, distances are non-metric \citep[i.e. the distance between A and B is not equal to the distance between B and A][]{Gerber2014-ol}.
And last but not least reason, In many cases, ordination might not be necessary.
For example, if an index (e.g. sum or ranges of variance) used to characterise disparity can use all of the data, it is not necessary to calculate it on the ordinated dataset \citep{Close2015-qi}.
For all of these reasons, multidimensional data analyses should not be ordinated automatically, and careful consideration should be given to whether the aim of the study can be achieved without ordination \citep{lloyd2016,lloyd2018}.

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.9\textwidth]{Figures/figure_space.pdf}
\caption{
    \tiny{
    Morphospaces: different mathematical representations of a morphospace.
    A trait matrix can be transformed into a distance matrix \citep[e.g in][]{Close2015-qi} or an ordinated matrix \citep[e.g. in][]{Brusatte2008-vx}.
    Here we consider all these matrices as being \textit{morphospaces}, i.e. objects containing all the combinations of traits and observations (albeit transformed differently).
    Visualisation: different  ways to represent the morphospace in 2D.
    Visualisations can use either trait plots (directly from the trait matrix); ``flattening'' ordination of the data (e.g. using non-metric multidimensional scaling forcing the dataset into two dimensions - note that variance is lost due to the distortion stress and non-metric properties of the resulting operation) ; or ordination axis plots (directly from the ordinated matrix).
    Note that in 2D representations, we consider it as good practice to plot both axis with the same scale (so to not visually distort the importance of one axis).
    }
    % @@@ Change references
}
\label{Fig:morphospace}
\end{figure}

\subsection{(b) Measuring disparity}
\label{section:metrics}

Most disparity datasets are multidimensional and, consequently, a large component of any disparity analysis involves considering how to extract a meaningful (i.e. interpretable) summary of disparity.
This is usually achieved with a disparity measure or index \citep{Hopkins2017-cf}.
As with any summary of multidimensional data, disparity indices will reflect only some aspects of the morphological variation, never its whole complexity \citep{GuillermeMOMS}.
It is therefore often beneficial to use more than one index to summarise different aspects of variation, guided by the aim of the study.

When considering only one dimension, disparity indices can be used to reflect the spread of the distribution (e.g. the range, quantiles or variance) or its central tendency (i.e., mean, median or mode).
Among these indices, some will have more attractive properties than others, such as sensitivity to outliers. Range and mean are highly sensitive, whereas quantiles, variance and median are less so, making them more or less appropriate for different questions.
For example, if we want to characterise the extent of morphospace occupied by a group (e.g. does group A occupy as much space as group B?), indices related to the spread of the group in the morphospace are most appropriate (e.g. volume
\citealt{Diaz2016-mr}, distance from the centroid \citealt{Hopkins2017-cf, Finlay2015-ft}, variance and range \citealt{Brusatte2008-fx}).
Furthermore, different aspects than variation (\textit{sensu} disparity) can be of interest: if we wish to describe the ``position'' of a group in a morphospace (e.g. does group A occupy the same space as group B?), indices related to the distance between the elements within a group and a fixed point in the morphospace are most appropriate \citep{GuillermeMOMS}.
Finally, if we aim to characterise the density of morphospace occupation (e.g. is group A more closely packed than group B?) indices related to the pairwise distances between elements will be most appropriate (e.g. nearest neighbour distance, pairwise distances, etc. \citealt{Close2015-qi} - see section \ref{section:ordination} above).

In addition to considering what properties of disparity indices should capture, it is important to also consider the mathematical properties of the indices and their associated caveats \citep{Wills2001-wh, Ciampaglio2001-iz}.
For example, measuring the sum of variance-covariance for each dimension of the space is equivalent of measuring the sum of variance in a PCA (since the co-variance is null).
However, this is not true of other transformations of the space or when not all dimensions or elements are considered, even in a PCA \citep{Legendre2012-va}.

Furthermore, multidimensional spaces also have some counter-intuitive properties that need to be considered such as the ``curse of dimensionality'' \citep{Bellman1966-mc}.
In spaces with some axis of variance lower than 1, product-based indices used as proxies of volumes (e.g. product of ranges, hypervolume, hypercube, etc.) can tend towards zero fairly quickly for spaces with even a modest number of dimensions \citep{Bellman1966-mc}.
Some other types of indices are also extremely sensitive to outliers and can be biased easily by sample size, for example range \citep{Foote1992-zs} or convex hull based indices \citep{Jackson2011-kq}.

\subsection{(c) Testing hypotheses on the evolution of disparity}
\label{section:testing}

No matter which disparity indices have been calculated, the research question must be framed in an appropriate statistical context.
The multidimensional statistical toolkit for ecology and evolution has been greatly expanded in recent years \citep{clavel2015mvmorph, Adams2018-mg}, but some of these advances have yet to be implemented in disparity analyses. Instead, hypothesis testing has mostly been confined to a small set of well-established methods.
One commonly used test is the non-parametric permutation analysis of variance \citep{Anderson2001-qb, Anderson2013-zt}, an analysis of variance (ANOVA) of the pairwise distances between different groups.
Although statistically valid, this test is not always directly related to the hypothesis under question.
For example, PERMANOVA tests whether two groups share the same variance/covariance based on a distance matrix between observations and therefore does not require the morphospace to be ordinated (see section \ref{section:ordination} above).
Statistical tests should be employed that are tailored to the question at hand, rather than simply following common practices.

It is also important to consider which data should be subjected to a statistical test.
For example, in morphological disparity analysis, especially for palaeobiological questions, data are often bootstrapped.
This has two advantages: (i) when the disparity index is unidimensional (e.g. the sum of variances), bootstrapping the data generates a distribution of the index that can be analysed using the vast statistical toolkit available for comparing distributions; (ii) when data are scarce, bootstrapping the data allows users to introduce variance, rendering the test less sensitive to outliers.
However, bootstrapped data are pseudoreplicates and thus non-independent.
This violates the assumptions of most parametric statistical tests.
Furthermore, the number of bootstrap pseudoreplicates will inevitably increase the false positive rate (Type I error).
These factors are often ignored in disparity analyses.
% VW: Surely there's a reference for this?
% MH: I might be wrong about this, but my impression is that CIs generated from bootstrapping are compared visually (like to assess significance in changes in disparity over time), but that statistical tests are not commonly applied.
Finally, it is important to understand the limitations of the dataset for performing statistical analysis.
Mainly, disparity analysis should be restrained to groups within the same morphospace and are more difficult between different morphospaces.
This can be the case when comparing elements with different numbers of landmarks or different landmark configurations for example: they will result in different morphospaces and comparing disparity indices between them is not trivial.

%TG: Add section d) within c)
\subsection{(d) Disparity and phylogeny}
\label{section:phylo}

As with all comparative datasets, the data used in disparity analyses are not independent because close relatives will tend to have more similar morphologies than more distant relatives \citep{Harvey1998-xg}.
Thus, for disparity analyses that consider groups with phylogenetic relationships (which is common) the non-independence between observations should be taken into account to correctly measure phenomenons.
However, it has been noted that some popular phylogenetic correction methods (like pPCA) can be inappropriate
% SG: In what way?
if incorrect assumptions are made about the data \citep{Uyeda2015}.
% @@@ref: https://academic.oup.com/sysbio/article/64/4/677/1649888
% SG: Which assumption? This part (e) seems a bit vague to me
% AG: as long as you are using most or ideally all of the variance in a dataset, the rotation of pPCA should have little to no impact on disparity analyses. polly et al 2013 i think

        % provide a thorough review of multivariate phylogenetic comparative methods, and so we do not consider them further here.
        % % AG: Given this is given quite short treatment, i think if you need room, you could cut it altogether or just have a sentence in another section stating this even more briefly

        % %TG: Add section e) within c)???
        % \subsection{(e) Ancestral state estimation in disparity-through-time analyses}
        % \label{section:ACE}

@@

\section{Disparity analyses for the future}
\label{section:future}

@@

\subsection{Appendix: morphospace definition in this paper}
\label{section:appendixmorphospace}
